---
title: "Measurement and DAGs"
date: "2026-02-05"
lightbox: auto
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))
source(here::here("R", "youtube-playlist.R"))

playlist_id <- "PLS6tnpTr39sFVrHdTifzlyS9dsJjhEhsF"

slide_details <- tibble::tribble(
  ~youtube_id,   ~title, ~slide, ~active,
  "V9zIL7bI9WI", "Introduction", "1", TRUE,
  "QkFhHOoYq-I", "Abstraction, stretching, and validity", "abstraction", FALSE,
  "kJQuH3nssfI", "Causal models", "causal-models", FALSE,
  "_qs_1B4ySWY", "Paths, doors, and adjustment", "paths-doors-adjustment", FALSE
)
```

## Readings

### Measurement

- {{< fa brands youtube >}} [The witch trial scene from *Monty Python and the Holy Grail*](https://www.youtube.com/watch?v=yp_l5ntikaU)
- {{< fa book >}} Chapter 5 in *Evaluation: A Systematic Approach* [@RossiLipseyHenry:2019]. This is available on iCollege.
- {{< fa book >}} Chapter 5 in *The Effect* [@Huntington-Klein:2021]

### DAGs

- {{< fa file-pdf >}} Julia M. Rohrer, "Thinking Clearly About Correlations and Causation: Graphical Causal Models for Observational Data" [@Rohrer:2018]. This will be posted on iCollege.
- {{< fa file-pdf >}} Lucy D’Agostino McGowan, Travis Gerke, and Malcolm Barrett, ["Causal inference is not just a statistics problem"](https://doi.org/10.1080/26939169.2023.2276446) [@DAgostinoMcGowanGerkeBarrett:2024]
- {{< fa file-pdf >}} Zachary M. Laubach et al., ["A Biologist’s Guide to Model Selection and Causal Inference"](https://doi.org/10.1098/rspb.2020.2815) [@LaubachMurrayHoke:2021]. This will be posted on iCollege.
- {{< fa book >}} Chapters 6 and 7 in *The Effect* [@Huntington-Klein:2021]

### DAG example page

- The [example page on DAGs](/example/dags.qmd) shows how to draw and analyze DAGs with both dagitty.net and R + {ggdag}


## Slides

The slides for today's lesson are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also click in the slides below and navigate through them with your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/04-slides")
slide_frame("/slides/04-slides.html")
```

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of special slide-specific commands.
:::


## Videos

```{r show-youtube-list, echo=FALSE, results="asis"}
youtube_list(slide_details, playlist_id)
```


## In-class stuff

Here are all the materials we'll use in class:

- [**Session 4 FAQ slides**](/slides/04-class.html) ([PDF](/slides/04-class.pdf))
- [Statistical vs. substantive significance](/news/2026-02-03_faqs_week-3.qmd#whats-the-difference-between-statistical-significance-and-substantial-significance)
- [Statistical testing in null worlds](https://nullworlds.andrewheiss.com/)
- Nick Huntington‐Klein et al., “The Influence of Hidden Researcher Decisions in Applied Microeconomics,” *Economic Inquiry* 59, no. 3 (2021): 944–60, <https://doi.org/10.1111/ecin.12992>.
- [Complex DAG example](https://stats.andrewheiss.com/testy-turtle/notebook/causal-model.html)
- [Dolly Parton's Imagination Library](https://imaginationlibrary.com/) + [video intro](https://www.youtube.com/watch?v=tscb01jfukc)
- [A dubious DAG](https://www.smithsonianmag.com/smart-news/growing-surrounded-books-may-bolster-skills-later-life-180970523/)

### Confounding, unblocked

```{=html}
<div class="ratio ratio-16x9">
<video controls loop>
<source src="/slides/img/05-class/video-confounding-unblocked.mp4" type="video/mp4">
</video>
</div>
```

### Confounding, blocked

```{=html}
<div class="ratio ratio-16x9">
<video controls loop>
<source src="/slides/img/05-class/video-confounding-blocked.mp4" type="video/mp4">
</video>
</div>
```

### Mediation

```{=html}
<div class="ratio ratio-16x9">
<video controls loop>
<source src="/slides/img/05-class/video-mediating.mp4" type="video/mp4">
</video>
</div>
```

### Collider bias

```{=html}
<div class="ratio ratio-16x9">
<video controls loop>
<source src="/slides/img/05-class/video-collider.mp4" type="video/mp4">
</video>
</div>
```

```{verbatim include=FALSE}
Playlist: https://www.youtube.com/playlist?list=PLS6tnpTr39sFVrHdTifzlyS9dsJjhEhsF

https://youtu.be/V9zIL7bI9WI
PMAP 8521 • (4) Measurement and DAGs: (0) Introduction

Introduction to the class session on outcome measurement and DAGs

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/04-content/


https://youtu.be/QkFhHOoYq-I
PMAP 8521 • (4) Measurement and DAGs: (1) Abstraction, stretching, and validity

Discussion of how to measure abstract concepts like program outcomes using a ladder of abstraction

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/04-content/

  
https://youtu.be/kJQuH3nssfI
PMAP 8521 • (4) Measurement and DAGs: (2) Causal models

Introduction to causal models and DAGs

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/04-content/


https://youtu.be/_qs_1B4ySWY
PMAP 8521 • (4) Measurement and DAGs: (3) Paths, doors, and adjustment

Discussion of how to identify associations, close backdoors, and isolate pathways in a DAG

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/04-content/

  
https://youtu.be/3euqrnD9w7c
PMAP 8521 • Example: DAGs with Dagitty

Demonstration of how to create and analyze DAGs with dagitty.net

Download the data and see the polished code at https://evalf20.classes.andrewheiss.com/example/dags/


https://youtu.be/uoAjyyToUTE
PMAP 8521 • Example: DAGs with R and ggdag

Demonstration of how to create and analyze DAGs with R and the {ggdag} package

Download the data and see the polished code at https://evalf20.classes.andrewheiss.com/example/dags/
```
