---
title: "Regression discontinuity I"
date: "2026-03-26"
---

```{r slides-videos, echo=FALSE, include=FALSE}
source(here::here("R", "slide-things.R"))
source(here::here("R", "youtube-playlist.R"))

playlist_id <- "PLS6tnpTr39sGYRrLn6OEoI5kry5gvJvdz"

slide_details <- tibble::tribble(
  ~youtube_id,   ~title, ~slide, ~active,
  "EqSMG277wVI", "Introduction", "1", TRUE,
  "vBDNcPHAGfg", "Arbitrary cutoffs and causal inference", "arbitrary-cutoffs", FALSE,
  "ZijPafzc-wY", "Drawing lines and measuring gaps", "lines-gaps", FALSE,
  "Or-VUzCZIBs", "Main RDD concerns", "main-concerns", FALSE
)
```

## Readings

- {{< fa book >}} Chapter 6 in *Impact Evaluation in Practice* [@wb-eval]
- {{< fa book >}} Chapter 20 in *The Effect* [@Huntington-Klein:2021]

### Regression discontinuity

- The [example page on regression discontinuity](/example/rdd.qmd) shows how to use R to analyze and estimate causal effects with regression discontinuity


## Slides

The slides for today's lesson are available online as an HTML file. Use the buttons below to open the slides either as an interactive website or as a static PDF (for printing or storing for later). You can also click in the slides below and navigate through them with your left and right arrow keys.

```{r show-slide-tabs, echo=FALSE, results="asis"}
slide_buttons("/slides/10-slides")
slide_frame("/slides/10-slides.html")
```

:::{.callout-tip}
**Fun fact**: If you type <kbd>?</kbd> (or <kbd>shift</kbd> + <kbd>/</kbd>) while going through the slides, you can see a list of special slide-specific commands.
:::


## Videos

```{r show-youtube-list, echo=FALSE, results="asis"}
youtube_list(slide_details, playlist_id)
```

## In-class stuff

Here are all the materials we'll use in class:

- [**Week 10 FAQ slides**](/slides/10-class.html) ([PDF](/slides/10-class.pdf))
- [Week 10 R code](/projects/week-10.zip) (on [Posit.cloud](https://posit.cloud/spaces/737332/content/12176556))

Quarto website resources:

- [Workshop on creating Quarto websites](https://andrewheiss.github.io/quarto-websites_2025-10/)
- [Creating a website](https://quarto.org/docs/websites/)
- [Bootswatch themes](https://bootswatch.com/)
- Publishing with [Posit Cloud Connect](https://quarto.org/docs/publishing/quarto-pub.html) ([see this too](https://andrewheiss-datavizsp26-11-exercise-recreate.share.connect.posit.cloud/instructions.html))
- Publishing with [Netlify Drop](https://app.netlify.com/drop)
- Publishing with [GitHub Pages](https://quarto.org/docs/publishing/github-pages.html)

Examples:

- This class website :)
- [My website](https://www.andrewheiss.com) ([`source`](https://github.com/andrewheiss/ath-quarto))
- Project-specific websites:
  - [Navigating Hostility](https://stats.andrewheiss.com/silent-skywalk/) ([`source`](https://github.com/andrewheiss/silent-skywalk))
  - [Double RDD with 990 data](https://stats.andrewheiss.com/motherly-dolphin/) ([`source`](https://github.com/andrewheiss/motherly-dolphin))
  - [Pandemic Pass](https://stats.andrewheiss.com/mountainous-mackerel/) ([`source`](https://github.com/andrewheiss/mountainous-mackerel))
  - [Derogations, Democratic Backsliding, and International Human Rights During the COVID-19 Pandemic](https://stats.andrewheiss.com/lemon-lucifer/) ([`source`](https://github.com/andrewheiss/lemon-lucifer))
- Interactive illustration website things: <https://www.andrewheiss.com/teaching/>
- Other websites:
  - [2024 AYS data visualization workshops](https://aysps.github.io/2024-aysps-dataviz/) ([`source`](https://github.com/aysps/2024-aysps-dataviz))
  - [Workshop on fancy regression](https://talks.andrewheiss.com/2024-11-13_udem-beyond-ols/) ([`source`](https://github.com/andrewheiss/2024-11-13_udem-beyond-ols))
  - [ForeignAssistance.gov backup](https://foreignassistance.andrewheiss.com/) ([`source`](https://github.com/andrewheiss/foreignassistance))

Other Quarto things

- [Citations](https://quarto.org/docs/authoring/citations.html) 
  - [Complete instructions](/resource/citations.qmd)
  - With [RStudio](https://posit.co/blog/rstudio-1-4-preview-citations/)
  - [Citation syntax](/resource/markdown.qmd#citations)
  - [Example of a citation-heavy document](https://stats.andrewheiss.com/snoopy-spring/) ([`source`](https://github.com/andrewheiss/snoopy-spring/blob/main/index.qmd))
- [Cross references](https://quarto.org/docs/authoring/cross-references.html)
- [Layout options](https://quarto.org/docs/authoring/article-layout.html)
- [Callouts](https://quarto.org/docs/authoring/callouts.html)
- Interactivity
  - [Plotly and `ggplotly()`](https://plotly.com/ggplot2/)
  - [Dashboards](https://quarto.org/docs/dashboards/)
  - [Observable JS](https://quarto.org/docs/interactive/ojs/) + [Observable Plot](https://observablehq.com/plot/getting-started) (like ggplot for Javascript)
  - Examples: 
    - [Penguin dashboard](https://jjallaire.github.io/ojs-penguins-dashboard/)
    - [Book tracker](https://books.heiss.website/2025/)
    - [Example book dashboard](https://www.andrewheiss.com/blog/2024/01/12/diy-api-plumber-quarto-ojs/_book/dashboard.html)
    - [Interactive maps with OJS](https://www.andrewheiss.com/blog/2025/02/10/usaid-ojs-maps/)

`ggplotly()` example:

```{r}
#| label: ggplotly-example
#| warning: false
#| message: false

library(tidyverse)
library(plotly)
library(gapminder)

plot_thing <- gapminder |> 
  filter(year == 2007) |> 
  ggplot(aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point(aes(text = country)) +  # text is a special aesthetic for plotly labels
  scale_x_log10()

ggplotly(plot_thing)
```

OJS example:

::: {.callout-note}
#### R

This is R code!
:::

```{r}
# Make the gapminder data available to Observable JS
ojs_define(gapminder = gapminder)
```

::: {.callout-note}
#### Observable
This is NOT R code! This is [Observable JS](https://quarto.org/docs/interactive/ojs/) code!
:::

```{ojs}
viewof current_year = Inputs.range(
  [1952, 2007], 
  {value: 1952, step: 5, label: "Year:"}
)

// Rotate the data so that it works with OJS
gapminder_js = transpose(gapminder)

// Filter the data based on the selected year
gapminder_filtered = gapminder_js.filter(d => d.year == current_year)

// Plot this thing
Plot.plot({
  x: {type: "log"},
  marks: [
    Plot.dot(gapminder_filtered, {
        x: "gdpPercap", y: "lifeExp", fill: "continent", r: 6,
        channels: {
          Country: d => d.country
        },
        tip: true
      }
    )
  ]}
)
```


```{verbatim include=FALSE}
Playlist: https://www.youtube.com/playlist?list=PLS6tnpTr39sGYRrLn6OEoI5kry5gvJvdz

https://youtu.be/EqSMG277wVI
PMAP 8521 • (10) Regression discontinuity I: (0) Introduction

Introduction to the class session on regression discontinuity designs

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/10-content/


https://youtu.be/vBDNcPHAGfg
PMAP 8521 • (10) Regression discontinuity I: (1) Arbitrary cutoffs and causal inference

Discussion of how arbitrary rules and cutoffs let us generate pseudo-treatment and control groups and find causal effects in observational data

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/10-content/

  
https://youtu.be/ZijPafzc-wY
PMAP 8521 • (10) Regression discontinuity I: (2) Drawing lines and measuring gaps

Discussion of how to draw lines in different ways around a discontinuity and how that influences the size of the gap

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/10-content/
  
  
https://youtu.be/Or-VUzCZIBs
PMAP 8521 • (10) Regression discontinuity I: (3) Main RDD concerns

Discussion of the main assumptions for regression discontinuity design

Access the slides and other materials at the course website at https://evalf20.classes.andrewheiss.com/content/10-content/
  
  
https://www.youtube.com/playlist?list=PLS6tnpTr39sFCr8amjT2Ld_FE34i26a3S

https://youtu.be/8CayGMcJoLY
PMAP 8521 • Example: Regression discontinuity with R: 1: Getting started

Demonstration of how to create a new RStudio project and get ready to analyze regression discontinuity data

Download the data and see the polished code at https://evalf20.classes.andrewheiss.com/example/rdd/


https://youtu.be/ViDChprm2IY
PMAP 8521 • Example: Regression discontinuity with R: 2: Sharp or fuzzy?

Demonstration of how to determine if a cutoff is sharp or fuzzy

Download the data and see the polished code at https://evalf20.classes.andrewheiss.com/example/rdd/


https://youtu.be/hqPVIt-72fY
PMAP 8521 • Example: Regression discontinuity with R: 3: Checking for manipulation

Demonstration of how to check for manipulation in the running variable near the cutpoint with a McCrary density test

Download the data and see the polished code at https://evalf20.classes.andrewheiss.com/example/rdd/


https://youtu.be/yYqO22l_jgc
PMAP 8521 • Example: Regression discontinuity with R: 4: Finding discontinuity in outcome

Demonstration of how to plot a discontinuity in an outcome based on a running variable and a cutpoint

Download the data and see the polished code at https://evalf20.classes.andrewheiss.com/example/rdd/


https://youtu.be/Por8Rd9QuBk
PMAP 8521 • Example: Regression discontinuity with R: 5: Measuring the gap parametrically

Demonstration of how to measure the gap at a discontinuity using parametric methods

Download the data and see the polished code at https://evalf20.classes.andrewheiss.com/example/rdd/
  

https://youtu.be/lfBkAZLX_Z4
PMAP 8521 • Example: Regression discontinuity with R: 6: Measuring the gap nonparametrically

Demonstration of how to measure the gap at a discontinuity using nonparametric methods with rdrobust

Download the data and see the polished code at https://evalf20.classes.andrewheiss.com/example/rdd/
```
