---
title: "Making fake data"
subtitle: "Problem set 9 --- PMAP 8521, Spring 2026"
author: "YOUR NAME HERE"  # Replace with your actual name
date: "DATE GOES HERE"  # Replace with the date
date-format: "long"
format:
  html:
    toc: true
  typst:
    toc: true
  pdf:
    toc: true
  docx:
    toc: true

# To make the list of steps cross-referenceable, I'm hijacking the "definition" 
# environment (see https://quarto.org/docs/authoring/cross-references.html#theorems-and-proofs)
# This changes the "Definition" text to "Process"
crossref:
  def-title: "Process"
  def-prefix: "Process"
---

# Overview

Many MPA and MPP programs offer a brief math camp in the weeks before students begin their graduate degrees, with the hope that it will help students by more prepared in math-heavy classes like statistics and microeconomics. 

You're interested in evaluating the effectiveness of a hypothetical math camp program. **Does attending math camp cause higher grades in statistics classes?**

This program is not randomized and it's not mandatory---anyone can decide to sign up (or not!), which means you have selection bias and confounding to worry about.

You don't have any data for this, but that's okay! You can simulate some data and set up the infrastructure for answering this question later with real data.

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)        # For ggplot, mutate(), filter(), and friends
library(broom)            # For converting models to data frames
library(ggdag)            # For drawing DAGs
library(scales)           # For rescaling data with rescale()
library(truncnorm)        # For truncated normal distributions
library(parameters)       # For converting models to data frames
library(modelsummary)     # For showing side-by-side regression tables
library(marginaleffects)  # For nicer post-estimation work
library(tinytable)        # For nice tables

set.seed(1234)  # Make any random stuff be the same every time you run this

# Turn off the message that happens when you use group_by() and summarize()
options(dplyr.summarise.inform = FALSE)
```

# The general process

At the [guide to example data](https://evalsp26.classes.andrewheiss.com/example/synthetic-data.html#specific-examples) at the course website, I provide this overview of the general steps involved in generating synthetic data:

::: {#def-process}

### How to generate synthetic data

1. Draw a DAG that maps out how all the columns you care about are related.
2. Specify how those nodes are measured.
3. Specify the relationships between the nodes based on the DAG equations.
4. Generate random columns that stand alone. Generate related columns using regression math. Consider adding random noise. This is an entirely trial and error process until you get numbers that look good. Rely *heavily* on plots as you try different coefficients and parameters. Optionally rescale any columns that go out of reasonable bounds. If you rescale, you'll need to tinker with the coefficients you used since the final effects will also get rescaled.
5. Verify all relationships with plots and models.
6. Try it out!
7. Save the data.

:::


# Background details

In this section, I provide you with the content for steps 1–3 in @def-process. In your final project, you'll have to do those steps yourself.

## DAG

> @def-process, step 1: Draw a DAG that maps out how all the columns you care about are related

@fig-math-camp-dag shows a DAG with just four nodes. Students' GRE scores and undergraduate GPAs confound both the choice to enroll in math camp and final class grades. Additionally, undergraduate degrees help cause GRE scores.

```{r}
#| label: fig-math-camp-dag
#| fig-cap: DAG showing the effect of math camp on final grades, confounded by GPA and GRE
#| fig-alt: |
#|   Directed acyclic graph with four nodes. Camp (the exposure) is on the
#|   far left and grade (the outcome) is on the far right. GPA and GRE are
#|   positioned in the middle row. Arrows point from GPA to GRE, from GPA
#|   and GRE to both camp and grade, and from camp to grade. This layout
#|   shows that GPA and GRE confound the relationship between math camp
#|   attendance and final grades.

math_camp_dag <- dagify(
  grade ~ camp + gpa + gre,
  camp ~ gpa + gre,
  gre ~ gpa,
  outcome = "grade",
  exposure = "camp",
  coords = list(
    x = c(camp = 1, gpa = 2, gre = 3, grade = 4),
    y = c(camp = 1, gpa = 2, gre = 2, grade = 1)
  )
)

ggdag_status(math_camp_dag) +
  theme_dag() +
  guides(color = "none") # Turn off legend
```


## Node measurement

> @def-process, step 2: Specify how those nodes are measured

Here's how these columns are measured:

- **GPA**: Scale from 1.5 to 4.0 with two decimals (i.e. values like 3.54 are fine). Mostly clustered around 3.5, plus or minus some amount that you decide. You'll either want to scale down whatever values you create with `rescale()`, or use `rtruncnorm()` to constrain the values within the plausible range.
- **GRE**: Scale from 130 to 170 with no decimals (i.e. values like 145.32 are bad). Mostly clustered around 150, plus or minus some amount that you decide. You'll either want to scale down whatever values you create with `rescale()`, or use `rtruncnorm()` to constrain the values within the plausible range.
- **Math camp**: Binary variable. Roughly 50% of students enroll in math camp.
- **Final grade**: Final exam score in a student's statistics class. Scale from 0 to 100 with one decimal (i.e. values like 75.2 are fine), mostly clustered around 70 (it's a hard test), plus or minus some amount that you decide. Nobody really scores below 20, so you can set the minimum to 20 if you want with `rescale()`.


## Node relationships

> @def-process, step 3: Specify the relationships between the nodes based on the DAG equations

In the DAG, there are three sets of relationships, or models:

- **GPA is associated with GRE scores**: Every 1 point increase in GPA (i.e. moving from a 2.0 to a 3.0) increases GRE scores by 10, on average.
- **Math camp is associated with undergraduate GPA and GRE scores**: Math camp here is a binary variable, so you can't say things like "every 1 point increase in GRE scores changes math camp by X". Instead, you'll want to do something similar to [the example for creating an effect for adjustment-based approaches](https://evalsp26.classes.andrewheiss.com/example/synthetic-data-ipw.html): create a column named `camp_score`, have GPA and GRE *decrease* that score (since people are less likely to enroll in math camp if they have higher GPAs and GRE scores), rescale that math camp score column to be between 0 and 1 to create a `camp_probability` column, and then randomly assign people to math camp based on those probabilities. I don't care what the size of the GPA and GRE effects are (since they'll be scaled down and binary-ized anyway)---just make it so that the probability of math camp goes down as GPA and GPA go up.
- **Final grade is caused by math camp, undergraduate GPA, and GRE scores**: Every 1 point increase in GPA increases final test scores by 10 points, on average. Every 1 point increase in GRE scores increase final test scores by 0.5 points. *Participating in math camp causes a boost of 10 points.* That's the causal effect you're building in.


# Task 1: Generate data

> @rem-process, step 4: Generate random columns that stand alone. Generate related columns using regression math. Consider adding random noise. This is an entirely trial and error process until you get numbers that look good. Rely heavily on plots as you try different coefficients and parameters. Optionally rescale any columns that go out of reasonable bounds. If you rescale, you'll need to tinker with the coefficients you used since the final effects will also get rescaled.

Before trying to get these variables to be related to each other, it's a good idea to see what good plausible values might be. "Math camp" and "final grade" are endogenous and are determined by GPA and GRE, so they're harder to see on their own, but we can visualize plausible distributions for GPA and GRE.

Generate and visualize some random values for both GPA and GRE. The values should match what's described above:

- **GPA**: Scale from 1.5 to 4.0, mostly clustered around 3.5, plus or minus some amount that you decide
- **GRE**: Scale from 130 to 170, mostly clustered around 150, plus or minus some amount that you decide

```{r}
#| label: check-general-distributions-gpa-gre

# Do stuff here
```

Use all the relationships in @fig-math-camp-dag to generate data for 2500 students.

You'll probably want to rescale columns so you don't end up with GPAs like 6.3, etc. Rescaling distorts the causal effects you build in, so you probably won't do something like `10 * camp`---it'll likely need to be higher.

```{r}
#| label: create-data

# Do stuff here
```


# Task 2: Verify relationships

> @def-process, step 5: Verify all relationships with plots and models

Make sure the relationships look good. GPA and GRE should be related to each other. Higher values of GPA and GRE should decrease the likelihood of math camp. GPA, GRE, and math camp should all be positively related with final grades.

Insert as many chunks as you need to here.


## Task 3: Try it out!

> @def-process, step 6: Try it out!

See if you can find the effect. Run two models:

- `grade ~ camp`: This is the naive, unadjusted, correlation-is-not-causation estimate. It's wrong.
- `grade ~ camp + gpa + gre`: This is the adjusted estimate. Backdoors have been closed and the coefficient for `camp` should be the causal effect. It should be around 10.
- (If you want to be super cool, adjust for GPA and GRE with inverse probability weighting too to see if the ATE for `camp` still holds. This is entirely optional.)
- Include the results of the models in a side-by-side regression table

Insert as many chunks as you need to here.


## Task 4: Save the data

> @def-process, step 7: Save the data

Make a version of your fake data that removes all the intermediate columns you made. Save the final clean data as a CSV file with `write_csv()`.

Insert a chunk here to do that.
